% !TeX spellcheck = ru_RU
% !TEX root = main.tex

\section{Введение}
\label{intro}

Реляционное программирование~\cite{TRS} --- это привлекательный подход, основанный на представлении программ как отношений.
%is an attractive technique, based on the idea of constructing programs as relations.
В результате реляционные программы можно \enquote{запускать} в различных \enquote{направлениях}, что позволяет, например, симулировать обратимые вычисления.
%As a result, relational programs can be ``queried'' in various ``directions'', making it possible, for example, to simulate reversed execution.
Это интересно не только с теоретической точки зрения, но обладает и практической ценностью: некоторые задачи выглядят проще, если их рассматривать как запросы к некоторой реляционной спецификации~\cite{WillThesis}.
%Apart from being interesting from a purely theoretical standpoint, this approach may have a practical value: some problems look much simpler when considered as queries to some relational specification~\cite{WillThesis}.
Можно назвать некоторое количество примеров, которые подтверждают это наблюдение:
алгоритм проверки типов для просто типизированного $\lambda$-исчисления (и в то же время алгоритм вывода типов и проверки населенности термов);
интерпретатор, способный синтезировать \enquote{квайны}~--- программы, вычисляющиеся в самих себя;
сортировка списков, способная породить все перестановки; т.п.
%There are a number of appealing examples confirming this observation:
%a type checker for simply typed lambda calculus (and, at the same time,
%a type inferencer and solver for the inhabitation problem),
%an interpreter (capable of generating ``quines''~--- programs producing themselves as a result)~\cite{Untagged},
%list sorting (capable of producing all permutations), etc.

Многие языки логического программирования (например, Prolog, Mercury~\cite{MercuryFirstPaper} или Curry~\cite{CurryFirstPaper}) в некоторой степени могут считаться реляционными.
%Many logic programming languages, such as Prolog, Mercury~\cite{MercuryFirstPaper}, or Curry~\cite{CurryFirstPaper} to some extent can be considered relational.
Нами был выбран \miniKanren\footnote{\url{http://minikanren.org}} как модельный язык, потому что он был специально спроектирован как предметно-ориенти\-рованный язык, встраиваемый в Scheme/Racket.
%We have chosen \miniKanren\footnote{\url{http://minikanren.org}} as a model language, because it was specifically designed as a relational DSL, embedded in Scheme/Racket.
Он довольно минималистичен --- может быть реализован с помощью малого количества структур данных и комбинаторов~\cite{MicroKanren, MuKanrenNew} --- и поэтому, он уже встроен во многие другие языки программирования общего назначения, включая Scala, Haskell and Standard ML.
%Being rather a minimalistic language, which can be implemented with just a few data structures and combinators~\cite{MicroKanren, MuKanrenNew}, \miniKanren{} found its way into dozens of host languages, including Scala, Haskell and Standard ML.
Парадигму в основе miniKanren можно описать как \enquote{легковесное логическое программирование}\footnote{Детальное сравнение miniKanren и Prolog можно найти здесь: \url{http://minikanren.org/minikanren-and-prolog.html} (проверено: \DTMDate{2025-01-10}).}.
%The paradigm behind \miniKanren can be described as ``lightweight logic programming''\footnote{An in-depth comparison of \miniKanren and Prolog can be found here: \url{http://minikanren.org/minikanren-and-prolog.html}}.

В данной работе исследуется задача встраивания miniKanren в OCaml\footnote{\url{http://ocaml.org}} --- статически типизированный функциональный язык с богатой системой типов.
Статическая типизация приносит ряд улучшений.
Во-первых, типизация даёт некоторый уровень корректности, отсеивая патологические программы, выдающие патологические результаты.
В контексте реляционного программирования типизация также помогает интерпретировать результаты запросов.
Часто ответы на реляционные запросы содержат свободные переменные, которые можно заменять на произвольные значения.
В типизированном случае эти переменные становятся типизированными, что упрощая понимание ответов, особенно с большим количеством свободных переменных.
Во-вторых, некоторые программы на \miniKanren{} требуют реализации и использования специальных ограничений.
В слабо типизированном случае, где всё может быть чем угодно, некоторые символы и структуры данных могут просачиваться в нежелательные места~\cite{Untagged}.
Чтобы это предотвращать в miniKanren для Scheme были добавление дополнительные ограничения (``\lstinline|absent$^o$|``, ``\lstinline|symbol$^o$|'').
Эти ограничения важны для слабых динамических систем типов, так как они отсекают ответы во время исполнения.
В статически типизированных языках эта работа может быть переложена на систему типов (при условии, что программист реализует их правильно), что не только улучшит производительность, но и уменьшит количество примитивов miniKanren, которые требуется реализовать.

%This paper addresses the problem of embedding \miniKanren into OCaml\footnote{\url{http://ocaml.org}}~--- a statically-typed functional language with
%a rich type system. A statically-typed implementation would bring us a number of benefits. First, as always,
%we expect typing to provide a certain level of correctness guarantees, ruling out some pathological programs, which
%otherwise would provide pathological results. In the context of relational programming, however, typing would additionally
%help us to interpret the results of queries.
%Often an answer to a relational query contains a number of
%free variables, which are supposed to take arbitrary values. In the typed case these variables become typed,
%facilitating the understanding of the answers, especially those with multiple free variables.
%Next, a number of \miniKanren
%applications require additional constraints to be implemented. In the untyped setting, when everything can be anything,
%some symbols or data structures tend to percolate into undesirable contexts~\cite{Untagged}. In order to prevent this from happening, some
%auxiliary constraints (``\lstinline|absent$^o$|'', ``\lstinline|symbol$^o$|'', etc.) were introduced.
%These constraints play a role
%of a weak dynamic type system, cutting undesirable answers out at runtime. Conversely, in a typed language this work can be
%entrusted to the type checker (at the price of enforcing an end user to write properly typed specifications), not only improving the
%performance of the system but also reducing the number of constraints which have to be implemented. Finally, it is rather natural
%to expect better performance of a statically-typed implementation.

В работе представлена реализация\footnote{\url{https://github.com/PLTools/OCanren}} реляционных комбинаторов и синтаксических расширений для языка OCaml,
которая в том числе поддерживает ограничения-неравенства~\cite{CKanren}, и поддерживает оптимизации faster-miniKanren\footnote{\url{TODO}}.
%We present an implementation of a set of relational combinators and syntax extensions for
%OCaml\footnote{The source code of our implementation is accessible from \url{https://github.com/dboulytchev/OCanren}.},
%which, technically speaking, corresponds to $\mu$Kanren~\cite{MicroKanren} with disequality
%constraints~\cite{CKanren}.
The contribution of our work is as follows:

\begin{enumerate}
\item Реализованное встраивание позволяет программисту использовать статическую типизацию и вывод типов в реляционных спецификациях. В частности, ошибки типизации находятся во время компиляции и типы логических переменных выводятся из контекста.
%Our embedding allows an end user to enjoy strong static typing and type inference in relational
%specifications; in particular, all type errors are reported at compile time and the types for
%all logical variables are inferred from the context.

\item Реализация основана на полиморфной унификации, которая также как и полиморфное сравнение, может использоваться для произвольных типов.
Реализация полиморфной унификации использует небезопасные возможности компилятора OCaml и полагается на специфичные особенности представления значений в памяти.
При этом можно доказать, что это не нарушает корректность типов.
%Our implementation is based on the \emph{polymorphic unification}, which, like the polymorphic comparison,
%can be used for arbitrary types. The implementation of polymorphic unification uses unsafe features and
%relies on the intrinsic knowledge of the runtime representation of values; we show, however, that this does not
%compromise type safety.

\item We describe a uniform and scalable pattern for using types for relational programming, which
helps in converting typed data to and from the relational domain. With this pattern, only one
generic feature (``\lstinline|Functor|'') is needed, and thus virtually any generic
framework for OCaml can be used. Although being rather a pragmatic observation, this pattern, we
believe, would lead to a more regular and easy way to maintain relational specifications.

\item We provide a simplified way to integrate relational and functional code. Our approach utilizes
a well-known pattern~\cite{Unparsing, DoWeNeed} for variadic function implementation and makes it
possible to hide the reification\footnote{First mention, explain what it is} of the answers phase from an end user.
\end{enumerate}

The rest of the paper is organized as follows: in the next section we provide a short overview of the related
works. Then we briefly introduce \miniKanren in
its original form to establish some notions; we do not intend to describe the language in its full bloom (interested readers can
refer to~\cite{TRS}). In Section~\ref{sec:goals} we describe some basic constructs behind a \miniKanren implementation, this time
in OCaml. In Section~\ref{sec:unification} we discuss polymorphic unification, and show that unification with
triangular substitution respects typing. Then we present our approach to handle user-defined types by injecting them
into the logic domain, and describe a convenient generic programming pattern, which can be used to implement the conversions from/to logic
domain. We also consider a simple approach and a more elaborate and efficient tagless variant (see Section~\ref{sec:injection}).
Section~\ref{sec:reification} describes top-level primitives and addresses the problem of relational and functional code integration.
Then, in Section~\ref{sec:examples} we present a set of relational examples, written with the aid of our
library. Section~\ref{sec:evaluation} contains the results of a performance evaluation and a comparison of our implementations
with existing implementation for Scheme. The final section concludes.

The authors would like to express a deep gratitude to the anonymous rewievers for their numerous constructive comments, Michael Ballantyne, Greg Rosenblatt,
and the other attendees of the miniKanren Google Hangouts, who helped the authors in understanding the subtleties of the original miniKanren
implementation, Daniel Friedman for his remarks on the preliminary version of the paper, and William Byrd for all his help and support, which cannot be
overestimated.

