% !TEX TS-program = pdflatex
% !TeX spellcheck = en_US
% !TEX root = main.tex

\section{Discussion and Future Work}

We presented the results of our work on guideline-based synthesis of GUI layouts. The approach we take makes it possible to
automatically build GUI layouts which by construction comply with a designer-specified set of rules. The prototype implementation
we developed allows to synthesize the layouts for the real-word industrial GUI components w.r.t. the real-world guidelines in
appropriate time.

One interesting question which may arise is if the application of relational programming is essential for this problem to be solved. Indeed, the set of guidelines
describes a matching (or rewriting) system which, in principle, can be directly implemented without any use of relational techniques. We argue, however,
that in this case a whole piece of work on justification of the correctness of the solution would have been repeated anew. In our
case, the justification trivially follows from the completeness of the \textsc{miniKanren} search and refutational completeness of our solution. We also speculate
that such a solution would require reinventing of some implementation techniques to support nondeterminism and backtracking, which are already native to relational
programming. Finally, the duality between patterns over structure and relational goals (initially unexpected for us), to our opinion, witnesses, that relational
programming is a truly relevant technique for this problem. %, and we are not stretching an owl over a globe.

%We can also consider the task of getting rid of extra solver (\textsc{Z3}) as relevant; this would not only simplify the infrastructure of the system, but also allow to
%integrate the constraint resolution phase into the constraint synthesis, improving the performance of the whole system.
%, and finally provide the completeness
%of the synthesizer, which we strictly speaking do not yet have. To do this, we plan to develop a relational integer inequality solver with a binary representation
%of fixed-size numbers. Note that for satisfactory performance it will also be necessary to extend \OCanren{} with inequality operations for such numbers in the
%form of a new type of constraint.

